---
title: "Data Analysis: Trend in life expectancy at birth in Sub-Saharan Africa since 1960"
author: "Baguinebie Bazongo"
date: "17/11/2019"
output:
  pdf_document: default
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary of main findings

Life expectancy at birth is recorded for 264 countries over the period 1960-2017. Cambodia and Rwanda experienced the lowest life expectancy in 1997 and 1993 respectively (18.9 vs. 26.2), while Hong Kong SAR, China and Japan experienced the highest life expectancy in 2017 (84.7 vs. 84.1).

When we examine Sub-Saharan Africa trend, figure 1 below shows overall increase in life expectancy at birth over 1960-2017. However, Rwanda experienced suddenly decreased in life expectancy since 1987 to reach its lowest level in 1993, a year of civil war. Although this decrease, Rwanda experienced a rapid increase in life expectancy since 1994, and this exceeded that of Sub-Saharan Africa since 2002. To conclude, life expectancy in Sub-Saharan Africa has significantly increased since 1997, but Rwanda had a higher progress in life expectancy compared to overall countries of the region.  



```{r echo = F, results = 'hide', message=F, warning=F,fig.show = 'hide'}
##### Part 1. Import data

##### Install required package for data analysis

##### Check if recquired packages for data analysis are installed and install if not installed

##WDI package to download World Bank development indicators

if (!require("WDI",character.only = TRUE)) {
  install.packages("WDI",dep=TRUE)
}

#Tidyverse package for data manaipulation and data visualization

if (!require("tidyverse",character.only = TRUE)) {
  install.packages("tidyverse",dep=TRUE)
}

# package to generate PDF output
if (!require("tinytex",character.only = TRUE)) {
  install.packages("tinytex",dep=TRUE)
}

##### Download the data set (rquire internet!)
data<-WDI(country = "all",indicator = "SP.DYN.LE00.IN",start = 1960)

##### Part 2. Clean data

##### Check for variables types
str(data)

#Summarize data
summary(data)

# rename life expectency at birth

names(data)[3]<-"exp"

#Check for missing values

sum(is.na(data$exp))

# Add missing values per country to data set

data<-data %>% 
  group_by(country) %>% 
  mutate(nb_c = sum(is.na(exp),na.rm=TRUE))

# visualize number of missing values per country
data %>% 
  distinct(country, .keep_all=TRUE)%>%
  filter(nb_c>2)%>%
  ggplot(aes(x=reorder(country,nb_c), y=nb_c))+
  geom_bar(stat="identity")+
  coord_flip()+
  labs(x="Country", y="Number of missing values", fill="")

# More details on missing values
# Israel
data %>% 
  filter(country=="Israel" & is.na(exp))%>% 
  print(n = nrow(.))


# Greenland
data %>% 
  filter(country=="Greenland" & is.na(exp))%>% 
  print(n = nrow(.))


# Kosovo
data %>% 
  filter(country=="Kosovo" & is.na(exp))%>% 
  print(n = nrow(.))

# St. Martin (French part)
data %>% 
  filter(country=="St. Martin (French part)" & is.na(exp))%>% 
  print(n = nrow(.))

# Faroe Islands
data %>% 
  filter(country=="Faroe Islands" & is.na(exp))%>% 
  print(n = nrow(.))

# West Bank and Gaza
data %>% 
  filter(country=="West Bank and Gaza" & is.na(exp))%>% 
  print(n = nrow(.))

# Liechtenstein
data %>% 
  filter(country=="Liechtenstein" & is.na(exp))%>% 
  print(n = nrow(.))

# Bermuda
data %>% 
  filter(country=="Bermuda" & is.na(exp))%>% 
  print(n = nrow(.))

# Serbia
data %>% 
  filter(country=="Serbia" & is.na(exp))%>% 
  print(n = nrow(.))

# Curacao ****
data %>% 
  filter(country=="Curacao" & is.na(exp))%>% 
  print(n = nrow(.))

# Dominica ****
data %>% 
  filter(country=="Dominica" & is.na(exp))%>% 
  print(n = nrow(.))

# select countries that have no more than 40 missing values (Serbia value)
life<-data %>% 
  filter(nb_c<=40)


# Checking for outliers using quantile function

q<-quantile(life$exp,probs = c(0.025,0.975),na.rm=TRUE)
print(q)

# Identify country out of range

#Examine outliers (lowest values)
life %>% 
  filter(exp<q[1])%>% 
  arrange(exp)%>% 
  print(n = nrow(.))

#Examine outliers (higest values)
life %>% 
  filter(exp>q[2])%>% 
  arrange(-exp)%>% 
  print(n = nrow(.))

```



```{r pressure, echo=FALSE,fig.width=10, fig.height=7,fig.fullwidth=TRUE}
data %>%
  select(country,exp,year)%>%
  filter((country=="Sub-Saharan Africa"|country=="Rwanda")& year<max(year)-1)%>%
  ggplot()+
  geom_line(aes(x = year, y = exp,color=country),size=1)+
  ggtitle("Figure 1. Life expectancy at birth in Sub Saharan Africa since 1960")+
  labs(x="year", y="Life expectancy (years)")+
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10))+
  theme(plot.title = element_text(face = "bold"))
```
